---
title: "Cartographic Explorations of the OpenStreetMap Database"
author: "Timothée Giraud"
date: "27 novembre 2017"
output: 
  html_document:
    fig_height: 4.6
    fig_width: 8.3
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_depth: 2
---

<br>

> This document contains a few cartographic explorations of the OpenStreetMap database


# Downloading and Cleaning Data

## Downloading

We use the [`osmdata` package](https://github.com/ropensci/osmdata) from [rOpenSci](https://ropensci.org/) to download OSM (OpenStreetMap) data.

The first step is to define a bounding box (`q0`) around the area we want to explore (Paris city). 
Then each features will be downloaded using a key/value selection (e.g.  `key = 'leisure', value = 'park'` to download parks).

Results will be `osmdata` objects containing the bounding box, call to the API and `sf` objects (points, line, polygons, etc.). We only use the relevant `sf` objects. 


```{r, eval=FALSE}
library(osmdata)
library(sf)
library(units)

# define a bounding box
q0 <- opq(bbox = c(2.2247, 48.8188, 2.4611, 48.9019)) 

# extract Paris boundaries
q1 <- add_osm_feature(opq = q0, key = 'name', value = "Paris")
res1 <- osmdata_sf(q1)
paris <- st_geometry(res1$osm_multipolygons[1,])

# extract the Seine river
q2 <- add_osm_feature(opq = q0, key = 'name', value = "La Seine")
res2 <- osmdata_sf(q2)
seine1 <- st_geometry(res2$osm_multilines)
q2b <- add_osm_feature(opq = q0, key = 'name', 
                       value = "La Seine - Bras de la Monnaie")
res2b <- osmdata_sf(q2b)
seine2 <- st_geometry(res2b$osm_lines)

# extract Parks and Cemetaries
q3 <- add_osm_feature(opq = q0, key = 'leisure', value = "park")
res3 <- osmdata_sf(q3)
parc1 <- st_geometry(res3$osm_polygons)
parc2 <- st_geometry(res3$osm_multipolygons)
q4 <- add_osm_feature(opq = q0, key = 'landuse', value = "cemetery")
res4 <- osmdata_sf(q4)
parc3 <- st_geometry(res4$osm_polygons)

# extract Quartiers
q5 <- add_osm_feature(opq = q0, key = 'admin_level', value = "10")
res5 <- osmdata_sf(q5)
quartier <- res5$osm_multipolygons

# extract Bars & Pubs
q6 <- add_osm_feature(opq = q0, key = 'amenity', value = "bar")
res6 <- osmdata_sf(q6)
bar <- res6$osm_points
q7 <- add_osm_feature(opq = q0, key = 'amenity', value = "pub")
res7 <- osmdata_sf(q7)
pub <- res7$osm_points

# extract Metro stations
q8 <- add_osm_feature(opq = q0, key = 'station', value = "subway")
res8 <- osmdata_sf(q8)
metro <- res8$osm_points

# extract Restaurant
q9 <- add_osm_feature(opq = q0, key = 'amenity', value = "restaurant")
res9 <- osmdata_sf(q9)
resto <- res9$osm_points

```

## Cleaning 

These data have to be cleaned before their mapping. 

```{r, eval=FALSE}
# use Lambert 93 projection (the french cartographic projection) for all layers
parc1 <- st_transform(parc1, 2154)
parc2 <- st_transform(parc2, 2154)
parc3 <- st_transform(parc3, 2154)
paris <- st_transform(paris, 2154)
seine1 <- st_transform(seine1, 2154)
seine2 <- st_transform(seine2, 2154)
quartier <- st_transform(quartier, 2154)
bar <- st_transform(bar, 2154)
pub <- st_transform(pub, 2154)
metro <- st_transform(metro, 2154)
resto <- st_transform(resto, 2154)

# make layers pretty
## Parcs and cemetaries are merged into a single layer, we only keep objects 
## greater than 1 ha
parc <- do.call(c, list(parc1, parc2, parc3))
parc <- st_union(x = st_buffer(parc,0), by_feature = F)
parc <- st_cast(parc, "POLYGON")
parc <- parc[st_area(parc)>=set_units(10000, "m^2")]
parc <- st_intersection(x = parc, y = paris)

## We only keep the part of the river within Paris boundaries
seine <- st_intersection(x = seine1, y = paris)
seine <- c(st_cast(seine[1])[2:5], seine[2])
seine <-c(seine, seine2)

## We only keep the bars and pubs within Paris boundaries
bar <- bar[!is.na(bar$name),]
pub <- pub[!is.na(pub$name),]
bars <- c(st_geometry(bar), st_geometry(pub))
bars <- st_intersection(x = bars, y = paris)

## We only keep the Paris quartiers
quartier <- quartier[substr(quartier$ref.INSEE,1,2)==75,]

## We only keep subway station within Paris boundaries
metro <- st_intersection(x = metro, y = paris)
metro <- metro[metro$station%in%"subway", ]
metro <- metro[metro$railway%in%"station",]
metro <- metro[metro$STIF.zone%in%1,]

## We only keep restaurants within Paris boundaries
resto <- resto[resto$amenity%in%"restaurant",]
resto <- resto[!is.na(resto$name),]
resto <- resto[,"cuisine"]
resto <- st_intersection(x = resto, y = paris)
resto$cuisine <- as.character(resto$cuisine)

# create a data folder
dir.create("data")
save(list= c("paris", "quartier", "seine", "parc", "bars", "metro", "resto"), 
     file = "data/paname.RData", compress = "xz")
```


>**If you want to skip the downloading & cleaning part, you can start here.** 

```{r, echo=TRUE, message=FALSE}
library(sf)
library(units)
load("data/paname.RData")
```

Let's visualize our data: 

```{r, message=FALSE}
library(cartography)

par(mar = c(0,0,1.2,0))
plot(paris, col = "#D9D0C9", border = NA, bg = "#FBEDDA")
plot(parc, col = "#CDEBB2", border = NA, add=T)
plot(seine, col = "#AAD3DF", add=T, lwd = 4)
plot(st_geometry(quartier), col = NA,lty = 2, lwd = 0.2, add=T)
plot(st_geometry(metro), add=T, col = "#005050", pch = 20, cex = 1)
plot(bars, add=T, col = "#0000ff", pch = 20, cex = 0.2)
plot(st_geometry(resto), add=T, col = "#ff0000", pch = 20, cex = 0.2)
plot(bars, add=T, col = "#0000ff", pch = 20, cex = 0.2)
plot(paris, add=T, lwd = 0.7)
legend(x = "right", legend = c("subway stations", "restaurants", "bars"), 
       col = c("#005050", "#ff0000", "#0000ff"), 
       pch = 20, pt.cex = c(1,0.2,0.2), cex = 0.7, bty = 'n')
layoutLayer(title = "Paris", scale = 1,
            north = T,
            tabtitle = TRUE, frame = FALSE,
            author = "Map data © OpenStreetMap contributors, under CC BY SA.", 
            sources = "cartography 2.0.2") 
```



# Bars in Paris

## Maps at the Quartier Level

We will first have a look at the number of bars in each quartier (a quartier is a neighborhood). 

```{r, message=FALSE, cache=TRUE, warning=FALSE}
# count the number of bars in each quartier
quartier$nbar <- lengths(st_covers(quartier, bars))
quartier$dbar <- quartier$nbar / set_units(st_area(quartier), "km^2")

par(mar = c(0,0,1.2,0))
plot(paris, col = "#D9D0C9", border = NA, bg = "#FBEDDA")
plot(parc, col = "#CDEBB2", border = NA, add=T)
plot(seine, col = "#AAD3DF", add=T, lwd = 4)
plot(st_geometry(quartier), col = NA,lty = 2, lwd = 0.2, add=T)
propSymbolsLayer(x = quartier, var = "nbar", inches = 0.2, col = "red", 
                 legend.pos = "topright", 
                 legend.title.txt = "Number of bars")
plot(paris, add=T, lwd = 0.7)
layoutLayer(title = "Spatial Distribution of Bars in Paris", scale = 1,
            tabtitle = TRUE, frame = F, 
            author = "Map data © OpenStreetMap contributors, under CC BY SA.", 
            sources = "cartography 2.0.2") 
```

We can also visualize the density per quartier.

```{r, message=FALSE, cache=TRUE, warning=FALSE}
par(mar = c(0,0,1.2,0))
plot(paris, col = "#D9D0C9", border = NA, bg = "#FBEDDA")
choroLayer(quartier, var = "dbar", border = NA, 
           method = "quantile", nclass = 6,
           col = carto.pal("wine.pal", 6),
           legend.pos = "topright",
           legend.title.txt = "Bars bensity\nper km2", add = TRUE)
plot(parc, col = "#E2CCB5", border = NA, add=T)
plot(seine, col = "#AAD3DF", add=T, lwd = 4)
plot(st_geometry(quartier), col = NA,lty = 2, lwd = 0.2, add=T)
plot(paris, add=T, lwd = 0.7)
layoutLayer(title = "Spatial Distribution of Bars in Paris", scale = 1,
            tabtitle = TRUE, frame = F, 
            author = "Map data © OpenStreetMap contributors, under CC BY SA.", 
            sources = "cartography 2.0.2") 
```

It is possible to combine the number of bars and their density in a single map. 

```{r, message=FALSE, cache=TRUE, warning=FALSE}
par(mar = c(0,0,1.2,0))
plot(paris, col = "#D9D0C9", border = NA, bg = "#FBEDDA")
plot(parc, col = "#CDEBB2", border = NA, add=T)
plot(seine, col = "#AAD3DF", add=T, lwd = 4)
plot(st_geometry(quartier), col = NA,lty = 2, lwd = 0.2, add=T)
propSymbolsChoroLayer(x = quartier, var = "nbar", var2 = "dbar",
                      inches = 0.2,
                      method = "quantile", nclass = 6,
                      col = carto.pal("wine.pal", 6),
                 legend.var2.pos  = "topright", 
                 legend.var2.title.txt = "Bars bensity\nper km2",
                 legend.var.title.txt = "Number of bars")
plot(paris, add=T, lwd = 0.7)
layoutLayer(title = "Spatial Distribution of Bars in Paris", scale = 1,
            tabtitle = TRUE, frame = F, 
            author = "Map data © OpenStreetMap contributors, under CC BY SA.", 
            sources = "cartography 2.0.2") 
```

## Maps at the Subway Station Level

Here, we first compute the mean minimal distance between two subway stations. 

```{r, eval=TRUE}
# distance between all stations
mat <- st_distance(metro, metro)
diag(mat) <- 10000
# distance to the nearest station
dmin <- apply(mat, 2, min)
# mean minimum distance 
(meandmin <- mean(dmin))
```

Then, we use this distance as a buffer around each station and count the number of bars in it. 

```{r, eval=TRUE, cache=TRUE}

# count the number of bars within this distance for each subway station
metro$nbar <- lengths(st_covers(st_buffer(x = metro, dist = meandmin), bars))

par(mar = c(0,0,1.2,0))
plot(paris, col = "#D9D0C9", border = NA, bg = "#FBEDDA")
plot(parc, col = "#CDEBB2", border = NA, add=T)
plot(seine, col = "#AAD3DF", add=T, lwd = 4)
plot(st_geometry(quartier), col = NA,lty = 2, lwd = 0.2, add=T)
plot(paris, add=T, lwd = 0.7)
propSymbolsChoroLayer(x = metro, var = "nbar", var2 = "nbar",
                      col = carto.pal("wine.pal",5), inches = 0.1,
                      method = "fisher-jenks", nclass = 5, add=T,
                      legend.var.pos = "topright", border = "grey50",
                      legend.var2.values.rnd = 0,
                      legend.var2.pos = "right",legend.var.style = "e",
                      legend.var.title.txt = "Number of bars\naround the station", 
                      legend.var2.title.txt = "Number of bars\naround the station")
layoutLayer(title = "Spatial Distribution of Bars in Paris", scale = 1,
            tabtitle = TRUE, frame = F, 
            author = "Map data © OpenStreetMap contributors, under CC BY SA.", 
            sources = "cartography 2.0.2") 
```




## Interactive Bar Map

Thanks to the [`leaflet`](https://github.com/rstudio/leaflet) package we can make an interactive map of the number of bars around each subway station. 


```{r, eval=TRUE}
library(leaflet)
# transform to lon/lat projection WGS84
metrounp <- st_transform(metro, 4326)
# order the table to have greater circles below smaller ones
metrounp <- metrounp[order(metrounp$nbar, decreasing = T),]
# compute the radii of the circles
metrounp$size <- sqrt(metrounp$nbar*15 / pi)
# create a label
metrounp$label <- paste0("<b>", metrounp$name, "</b> <br>", 
                         metrounp$nbar," bars à proximité.")
# create the interactive map...
m <- leaflet(padding = 0)
m <- addTiles(m)
m <- addCircleMarkers(map = m, 
                      lng = st_coordinates(metrounp)[,1], 
                      lat = st_coordinates(metrounp)[,2], 
                      radius = metrounp$size, weight = 0.25, 
                      stroke = T, opacity = 100,
                      fill = T, fillColor = "#920000", 
                      fillOpacity = 100,
                      popup = metrounp$label,
                      color = "white")
# ... and display it
m
```


## Smoothed analysis of the bars spatial distribution

With the [`spatstat`](https://github.com/spatstat/spatstat) package we compute a smoothed image of the spatial distribution of bars. We use a Spatial Kernel Density Estimation with a gaussian smoothing kernel. 

```{r, message=FALSE, cache=TRUE, warning=FALSE}
library(spatstat)
library(maptools)
library(raster)
bb <- as(bars, "Spatial")
bbowin <- as.owin(as(paris, "Spatial"))
pts <- coordinates(bb)
p <- ppp(pts[,1], pts[,2], window=bbowin)
ds <- density.ppp(p, sigma = 200, eps = c(20,20), kernel="epanechnikov")
rasdens <- raster(ds) * 1000 * 1000
rasdens <- rasdens+1
par(mar = c(0,0,1.2,0))
bks <- getBreaks(values(rasdens), nclass = 12, method = "equal")
cols <- colorRampPalette(c("black","#940000", "white"))(length(bks)-1)
plot(paris, col = NA, border = NA, main="", bg = "#FBEDDA")
plot(rasdens, breaks= bks, col=cols, add = T,legend=F)
legendChoro(pos = "topright",cex = 0.7,
            title.txt = "Bar density\nKDE, sigma=200m\n(bars per km2)",
            breaks = bks-1, nodata = FALSE,values.rnd = 0,
            col = cols)
plot(seine, col = "#AAD3DF", add=T, lwd = 4)
plot(parc, col = "#CDEBB235", border = NA, add=T)
plot(quartier$geometry, add=T, border = "white",lty = 2, lwd = 0.1)
plot(bars, add=T, col = "#0000ff90", pch = 20, cex = 0.1)
plot(paris, col = NA, add=T)
north(pos = c(661171.8,6858500))
layoutLayer(title = "Spatial Distribution of Bars in Paris", scale = 1,
            tabtitle = TRUE, frame = FALSE,
            author = "Map data © OpenStreetMap contributors, under CC BY SA.", 
            sources = "cartography 2.0.2") 
```




# Restaurants in Paris

We can make the same kind of maps using other OSM points data. The following are maps of restaurants. Since we would have to repeat almost the same code multiple times, we create a custom function to create these maps. 

## Creating a Function
```{r, message=FALSE, cache=TRUE, warning=FALSE}
densityMap <- function(x = resto, cuisine = NULL, title, sigma ){
  opar <- par(mar = c(0,0,0,0))
  if(!is.null(cuisine)){
    x <- x[x$cuisine %in% cuisine, ]
  }
  bb <- as(x, "Spatial")
  bbowin <- as.owin(as(paris, "Spatial"))
  pts <- coordinates(bb)
  p <- ppp(pts[,1], pts[,2], window=bbowin)
  ds <- density.ppp(p, sigma = sigma, eps = c(20,20))
  rasdens <- raster(ds) * 1000 * 1000
  rasdens <- rasdens+1
  bks <- getBreaks(values(rasdens), nclass = 12, method = "equal")
  cols <- colorRampPalette(c("black", "#940000", "white"))(length(bks)-1)
  plot(paris, col = NA, border = NA, main="")
  image(rasdens, breaks= bks, col=cols, add = T,legend=F)
  legendChoro(pos = "topright",cex = 0.7,title.cex = .8,
              title.txt = paste0(title, "\nsigma=500m,\nn=",nrow(pts)),
              breaks = bks-1, nodata = FALSE,values.rnd = 0,
              col = cols)
  plot(seine, col = "#AAD3DF", add=T, lwd = 4)
  plot(parc, col = "#CDEBB235", border = NA, add=T)
  plot(quartier$geometry, add=T, border = "white",lty = 2, lwd = 0.1)
  plot(st_geometry(x), add=T, col = "blue", pch = 20, cex = 0.1)
  plot(paris, col = NA, add=T)
  barscale(size = 1)
  mtext(text = "cartography 2.0.2\nMap data © OpenStreetMap contributors, under CC BY SA.", 
        side = 1, line = -1, adj = c(0.01,0.01), cex = 0.6, font = 3)
  north(pos = c(661171.8,6858051))
  par(opar)
}
```

Here is a test of the function for all restaurants: 
```{r, message=FALSE, cache=TRUE, warning=FALSE}
densityMap(x = resto, title = "All Restaurants", sigma = 500)
```



## Restaurants Maps by Cuisine

```{r, message=FALSE, cache=TRUE, warning=FALSE}
densityMap(x = resto, cuisine = c('japanese','sushi'), 
           title = "Japanese cuisine", 
           sigma = 500)
densityMap(x = resto, cuisine = c('korean'), 
           title = "Korean cuisine", 
           sigma = 500)
densityMap(x = resto, cuisine = c('asian', "thai", "vietnamese", "chinese"), 
           title = "Other asian cuisine", 
           sigma = 500)
densityMap(x = resto, cuisine = c('indian'), 
           title = "Indian cuisine", 
           sigma = 500)
densityMap(x = resto, cuisine = c('pizza','italian'), 
           title = "Italian cuisine", 
           sigma = 500)
```

# Reproducibility Information

```{r}
sessionInfo()
```